---
title: "PERMANOVA species beta diversity by study group"
description: "Bray–Curtis dissimilarity on species relative abundance; group effect tested with permutational MANOVA."
date: 2026-05-19
categories: [permanova, beta-diversity, species, pre-specified]
author: "IBD Capstone team"
---

## Research question

Does **study group** (Non-IBD, Active IBD, Quiescent) explain differences in **species-level community composition** (beta diversity)?

**Null:** Centroids (and dispersion, under default `adonis2`) do not differ among study groups.

## Data

- **Response:** Species relative abundance (`s__*` columns after wide reshape).
- **Predictor:** `Study_group_new`.
- **Unit:** Stool sample (`Sample_ID`).
- **Input:** `data/processed/species.csv` (long format from `make save`; pivoted via `load_taxa_wide()`).
- **Covariates in this model:** none (add to formula if extended later).

## Methods

| Item | Choice |
|------|--------|
| Test | PERMANOVA (`vegan::adonis2`) |
| Distance | Bray–Curtis (`vegan::vegdist`, `method = "bray"`) |
| Formula | `dist_matrix ~ Study_group_new` |
| Permutations | 999 |
| Ordination (visualization) | PCoA (`stats::cmdscale`, k = 2) |
| Ellipses | 95% normal ellipses on PCoA (`ggplot2::stat_ellipse`) |

```{r}
#| label: setup
#| include: false
library(here)
library(knitr)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
here::i_am("posts/2026-05-19-permanova-species-beta-diversity/index.qmd")
source(here::here("R", "permanova_helpers.R"))

level <- "species"
cfg <- TAXA_LEVELS[[level]]
wide_df <- load_taxa_wide(level)
fit <- run_permanova_study_group(wide_df, cfg$prefix)
permanova_result <- fit$permanova
```

## Assumptions and diagnostics

- PERMANOVA is **permutation-based**; does not assume multivariate normality.
- **Homogeneity of dispersion:** not checked in this post; consider `betadisper()` + `permutest()` if interpretation hinges on location vs dispersion.
- **Pseudo-F** and **R²** are reported by `adonis2` for the grouping term.

## Results

```{r}
#| label: permanova-table
#| echo: false
knitr::kable(
  permanova_to_df(permanova_result),
  digits = 4,
  col.names = c("Term", "Df", "Sum of squares", "R²", "F", "Pr(>F)")
)
```

```{r}
#| label: results-inline
#| echo: false
#| results: asis
cat(permanova_summary_text(permanova_result, cfg$label))
```

## Figure

PCoA of Bray–Curtis distances, coloured by study group.

```{r}
#| label: fig-pcoa-permanova
#| fig-cap: "PCoA of species Bray–Curtis dissimilarity. Points are samples; dashed ellipses are 95% normal ellipses by study group."
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center

plot_pcoa_study_group(
  fit$dist,
  fit$metadata,
  permanova_result,
  cfg$label
)
```

```{r}
#| label: save-figure
#| include: false
dir.create(here("figures"), showWarnings = FALSE, recursive = TRUE)
ggsave(
  here("figures", "permanova-species-pcoa.png"),
  width = 7,
  height = 5.5,
  dpi = 300
)
```

## Interpretation

A significant PERMANOVA indicates **differences in community composition** among study groups in this distance-based framework. Follow-up may include pairwise PERMANOVA, dispersion tests, or differential abundance methods; those should be logged as separate posts.

## Reproducibility

- **Notebook:** `notebooks/permanova.Rmd`
- **Helpers:** `stats/R/permanova_helpers.R`
- **Pipeline:** `make wrangle` → `make save`
- **Render site:** `quarto render stats` from repository root
