---
title: "Kruskal–Wallis  alpha diversity by study group and fibre restriction"
description: "Non-parametric comparison of Shannon, richness, and other alpha metrics across study group and fibre-restriction levels."
date: 2026-05-19
categories: [alpha-diversity, pre-specified]
author: "IBD Capstone team"
---

## Research question

1. Do **alpha diversity metrics** differ across **study group**?
2. Do they differ across **fibre restriction** level (None, Low, Mid, High)?

**Null (each metric):** Distributions are equal across groups.

## Data

- **Response:** `Value` (alpha diversity metric).
- **Predictors:** `Study_group_new`, `Fiber_restriction`.
- **Unit:** Sample (one row per sample × metric in long format).
- **Input:** `data/intermediate/alpha_long.rds` (from `make wrangle`).
- **Pre-specified script:** `src/mycobiome/03b_diversity_stats.R`.

## Methods

| Item | Choice |
|------|--------|
| Test | Kruskal–Wallis (`rstatix::kruskal_test`) |
| Stratification | One test per `Diversity_metric` (`group_by(Diversity_metric)`) |
| Formulae | `Value ~ Study_group_new`; `Value ~ Fiber_restriction` |
| Post-hoc | Not run in pipeline (Dunn + BH commented in `stool_eda.Rmd`) |
| Significance | α = 0.05 (uncorrected across metrics) |

```{r}
#| label: setup
#| include: false
library(dplyr)
library(rstatix)
library(ggplot2)
library(tidyr)
library(here)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
# Without this, interactive sessions anchor to the .Rproj root and here::here("..") breaks.
here::i_am("posts/2026-05-19-alpha-diversity-kruskal-wallis/index.qmd")

alpha_long <- readRDS(here::here("..", "data", "intermediate", "alpha_long.rds"))
```

```{r}
#| label: run-tests
kw_diversity_group <- alpha_long |>
  group_by(Diversity_metric) |>
  kruskal_test(Value ~ Study_group_new)

kw_diversity_fibre <- alpha_long |>
  group_by(Diversity_metric) |>
  kruskal_test(Value ~ Fiber_restriction)
```

## Assumptions and diagnostics

- Kruskal–Wallis compares **distributions** (rank-based); does not assume normality.
- **Independence:** assumed for samples; repeated measures per participant not modelled here.
- **Multiple metrics:** several diversity indices tested; consider FDR if reporting many metrics together.

## Results

### By study group

```{r}
#| label: kw-group-table
#| echo: false
knitr::kable(kw_diversity_group, digits = 4)
```

### By fibre restriction

```{r}
#| label: kw-fibre-table
#| echo: false
knitr::kable(kw_diversity_fibre, digits = 4)
```

```{r}
#| label: sig-summary
#| echo: false
#| results: asis
sig_group <- kw_diversity_group |> filter(p < 0.05)
sig_fibre <- kw_diversity_fibre |> filter(p < 0.05)
cat("**Significant by study group (p < 0.05):** ",
    if (nrow(sig_group) == 0) "none" else paste(sig_group$Diversity_metric, collapse = ", "),
    "\n\n",
    "**Significant by fibre restriction (p < 0.05):** ",
    if (nrow(sig_fibre) == 0) "none" else paste(sig_fibre$Diversity_metric, collapse = ", "),
    sep = "")
```

## Figure

Boxplots of alpha diversity by study group (one panel per metric).

```{r}
#| label: fig-alpha-IBD-boxplots
#| fig-cap: "Alpha diversity by study group. Kruskal–Wallis p-values are in the results tables above."
#| fig-width: 8
#| fig-height: 6
#| fig-align: center

p_group <- ggplot(
  alpha_long,
  aes(x = Study_group_new, y = Value, fill = Study_group_new)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1) +
  facet_wrap(~ Diversity_metric, scales = "free_y") +
  labs(
    x = NULL,
    y = "Alpha diversity value",
    fill = "Study group"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
p_group
```

```{r}
#| label: fig-alpha-fibre-boxplots
#| fig-cap: "Alpha diversity by fibre restrictive diet. Kruskal–Wallis p-values are in the results tables above."
#| fig-width: 8
#| fig-height: 6
#| fig-align: center

p_fibre <- ggplot(
  alpha_long,
  aes(x = Fiber_restriction, y = Value, fill = Fiber_restriction)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1) +
  facet_wrap(~ Diversity_metric, scales = "free_y") +
  labs(
    x = NULL,
    y = "Alpha diversity value",
    fill = "Fibre Restriction",
    title = "Alpha diversity by fibre restricted diets"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
p_fibre
```

```{r}
#| label: save-figures
#| include: false
fig_dir <- here::here("figures")
ggsave(file.path(fig_dir, "alpha-diversity-kw-by-group.png"), plot = p_group, width = 8, height = 6, dpi = 300)
ggsave(file.path(fig_dir, "alpha-diversity-kw-by-fibre.png"), plot = p_fibre, width = 8, height = 6, dpi = 300)
```

## Interpretation

Significant Kruskal–Wallis results indicate **at least one group** differs in distribution for that metric; they do not identify which pairs differ. Use Dunn tests (with multiplicity adjustment) in a follow-up post if needed.

## Reproducibility

- **Notebook:** `notebooks/stool_eda.Rmd` (§ Inference)
- **Makefile target:** `make diversity-stats`
