---
title: "PCA and t-SNE Analysis"
output:
  html_document:
    df_print: paged
---

# Introduction

Purpose: Dimensionality reduction of the mycobiome using z-score normalised phylum, genus, and species abundance matrices combined into a single feature matrix.

PCA is used to summarise variance structure and identify the taxa that drive sample separation (loadings); t-SNE provides a non-linear complement to visualize local clustering. Both are rendered as interactive 3D plots coloured by study group (Active IBD / Quiescent / Non-IBD) to assess whether mycobiome composition separates by disease status.

This notebook is an artifact of the initial exploration of the data. Parts of the core pipeline were derived from this notebook; however, it does not reflect the current analysis. Please interpret with caution. It is included in the repo for informational purposes. The input data is derived from the core pipeline, and could be rerun manually to reflect current data.

# Libraries

```{r}
library(Rtsne)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(scales)
library(plotly)
```

# Load Data

```{r}
phylum <- readRDS("../data/intermediate/phylum_z.rds")
genus <- readRDS("../data/intermediate/genus_z.rds")
species <- readRDS("../data/intermediate/species_z.rds")
meta_data <- readRDS("../data/intermediate/meta_data.rds")
```

# Combine Phylum, Genus, and Species

```{r}
taxa_info <- bind_rows(
  data.frame(Taxonomic_Level = "Phylum", Taxon = rownames(phylum)),
  data.frame(Taxonomic_Level = "Genus", Taxon = rownames(genus)),
  data.frame(Taxonomic_Level = "Species", Taxon = rownames(species))
) |>
  mutate(Feature = paste(Taxonomic_Level, Taxon, sep = " | "))

taxa_df <- rbind(phylum, genus, species) |>
  as.data.frame(check.names = FALSE)

rownames(taxa_df) <- taxa_info$Feature

taxa_df |>
  rownames_to_column("Feature") |>
  head()
```

```{r}
X <- t(as.matrix(taxa_df))

sample_meta <- data.frame(Sample_ID = rownames(X)) |>
  left_join(meta_data, by = "Sample_ID")

```

# PCA

```{r}
pca <- prcomp(X, center = TRUE, scale. = TRUE)

pca_variance <- pca$sdev^2 / sum(pca$sdev^2)

pca_variance_df <- data.frame(
  PC = paste0("PC", seq_along(pca_variance)),
  Component = seq_along(pca_variance),
  Variance = pca_variance,
  Cumulative = cumsum(pca_variance)
)

pca_scores <- data.frame(
  Sample_ID = rownames(X),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  PC3 = pca$x[, 3]
) |>
  left_join(sample_meta, by = "Sample_ID")
```

## Cumulative Variance

```{r}
pca_variance_df |>
  filter(Component <= 10) |>
  mutate(
    Variance = round(Variance * 100, 2),
    Cumulative = round(Cumulative * 100, 2)
  )
```

```{r}
ggplot(
  pca_variance_df |> filter(Component <= 10),
  aes(x = Component, y = Cumulative)
) +
  geom_line(color = "#4E79A7") +
  geom_point(color = "#4E79A7", size = 2) +
  scale_x_continuous(breaks = 1:10, labels = paste0("PC", 1:10)) +
  scale_y_continuous(labels = scales::percent, limits = c(0,1)) +
  labs(
    title = "Cumulative PCA Variance",
    x = "Principal component",
    y = "Cumulative variance explained"
  ) +
  theme_minimal()
```

## Loadings

```{r}
pca_loadings <- as.data.frame(pca$rotation[, 1:3, drop = FALSE]) |>
  rownames_to_column("Feature") |>
  left_join(taxa_info, by = "Feature") |>
  pivot_longer(
    starts_with("PC"),
    names_to = "PC",
    values_to = "Loading"
  ) |>
  mutate(Abs_Loading = abs(Loading))

pca_loadings |>
  group_by(PC) |>
  slice_max(Abs_Loading, n = 10) |>
  arrange(PC, desc(Abs_Loading))
```

```{r}
ggplot(
  pca_loadings |>
    filter(PC %in% c("PC1", "PC2", "PC3")) |>
    group_by(PC) |>
    slice_max(Abs_Loading, n = 10) |>
    ungroup(),
  aes(x = Abs_Loading, y = reorder(Feature, Abs_Loading), fill = Loading > 0)
) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ PC, scales = "free_y") +
  scale_fill_manual(values = c("#F28E2B", "#4E79A7")) +
  labs(
    title = "Top PCA Loadings",
    x = "Absolute loading",
    y = NULL
  ) +
  theme_minimal()
```

## 3D PCA Figure

```{r}
pca_axis_labels <- paste0(
  c("PC1", "PC2", "PC3"),
  " (",
  round(pca_variance_df$Variance[1:3] * 100, 1),
  "%)"
)

plot_ly(
  pca_scores,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~Study_group_new,
  hovertext = ~paste("Sample:", Sample_ID, "<br>Study group:", Study_group_new),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 5, opacity = 0.85)
) |>
  layout(
    title = "3D PCA: Phylum + Genus + Species",
    scene = list(
      xaxis = list(title = pca_axis_labels[1]),
      yaxis = list(title = pca_axis_labels[2]),
      zaxis = list(title = pca_axis_labels[3])
    ),
    legend = list(title = list(text = "Study group"))
  )
```

# t-SNE

```{r}
set.seed(123)

tsne_perplexity <- 7

tsne <- Rtsne(
  X,
  dims = 3,
  perplexity = tsne_perplexity,
  pca = FALSE,
  check_duplicates = FALSE
)

tsne_scores <- data.frame(
  Sample_ID = rownames(X),
  tSNE1 = tsne$Y[, 1],
  tSNE2 = tsne$Y[, 2],
  tSNE3 = tsne$Y[, 3]
) |>
  left_join(sample_meta, by = "Sample_ID")
```

## 3D t-SNE Figure

```{r}
plot_ly(
  tsne_scores,
  x = ~tSNE1,
  y = ~tSNE2,
  z = ~tSNE3,
  color = ~Study_group_new,
# Purpose

Dimensionality reduction of the mycobiome using z-score normalised phylum,
genus, and species abundance matrices combined into a single feature matrix.
PCA is used to summarise variance structure and identify the taxa that drive
sample separation (loadings); t-SNE provides a non-linear complement to
visualise local clustering. Both are rendered as interactive 3D plots coloured
by study group (Active IBD / Quiescent / Non-IBD) to assess whether mycobiome
composition separates by disease status.

# Libraries = ~paste("Sample:", Sample_ID, "<br>Study group:", Study_group_new),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 5, opacity = 0.85)
) |>
  layout(
    title = paste0("3D t-SNE: Phylum + Genus + Species (perplexity = ", tsne_perplexity, ")"),
    scene = list(
      xaxis = list(title = "t-SNE 1"),
      yaxis = list(title = "t-SNE 2"),
      zaxis = list(title = "t-SNE 3")
    ),
    legend = list(title = list(text = "Study group"))
  )
```
