---
title: "PCA and clustering on combined diet and mycobiome profiles"
description: "Participant-level PCA of scaled nutrient intake and genus relative abundance, with hierarchical clustering to explore diet–mycobiome groupings."
date: 2026-05-29
categories: [pre-specified, pca, clustering, diet, genus]
author: "IBD Capstone team"
---

## Research question

Can **participants** be grouped by combined **diet (nutrients)** and **mycobiome (genus taxa)** profiles?

**Pre-specified method; exploratory interpretation:** with 12 participants, PCA and clustering describe patterns in the data but do not provide stable cluster validation (no formal cluster significance testing).

## Data

| Role | Source |
|------|--------|
| Genus taxa | `data/processed/genus.csv` (`make save`) |
| Diet | `data/processed/dietary_cleaned.xlsx` |
| Disease group (reference) | `Study_group_new` on mycobiome metadata |

- **Unit of analysis:** **Participant** (one row per `Participant_ID`).
- **Mycobiome:** Mean genus relative abundance across stool samples per participant.
- **Diet:** Mean nutrient intake across diary days per participant (same as `merge_files.R`).
- **Features:** Panel of nutrients from `NUTRIENT_FIELDS` plus all genus columns (`g__*`) with non-zero variance across participants.
- **Preprocessing:** Features are centered and scaled before PCA and clustering.

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(knitr)
library(here)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
here::i_am("posts/2026-05-29-diet-mycobiome-pca-clustering/index.qmd")
source(here::here("R", "pca_clustering_helpers.R"))

participant_df <- build_participant_feature_table()
feature_list <- prepare_clustering_features(participant_df)
pca_fit <- run_feature_pca(feature_list)
cluster_fit <- run_hclust_clusters(feature_list, k_values = c(2L, 3L))
scores <- add_clusters_to_scores(pca_fit$scores, cluster_fit)

n_participants <- nrow(scores)
n_features <- length(feature_list$features)
```

**Coverage:** `r n_participants` participants, `r n_features` features (`r feature_list$n_nutrients` nutrients + `r feature_list$n_taxa` genera).

## Methods

| Item | Choice |
|------|--------|
| Integration | Left-join participant-mean genus abundances and nutrients on `Participant_ID` |
| PCA | `stats::prcomp`, `center = TRUE`, `scale. = TRUE` |
| Clustering | Hierarchical (Ward.D2) on Euclidean distance of scaled features |
| Cluster sizes | k = 2 and k = 3 (`cutree`) |
| Reference labels | `Study_group_new` overlaid on PCA (not used to fit clusters) |
| Notebook reference | `notebooks/pca_tsne.Rmd` (taxa-only PCA/t-SNE) |

```{r}
#| label: run-analysis
# Results objects created in setup: pca_fit, cluster_fit, scores
```

## Assumptions and diagnostics

- **Sample size:** 12 participants; PCA axes and clusters are unstable and sensitive to outliers.
- **Scale:** Nutrients (g, kcal, mg) and compositional genus proportions are jointly z-scored; relative weighting is arbitrary but standard for integrated PCA.
- **Compositional data:** Genus abundances are not CLR-transformed here; results complement Bray–Curtis / PERMANOVA analyses on taxa alone.
- **Independence:** Technical stool replicates are averaged before analysis.

## Results

### Variance explained

```{r}
#| label: variance-table
#| echo: false
pca_fit$variance_df |>
  filter(Component <= min(10L, n_participants)) |>
  mutate(
    Variance = round(Variance * 100, 2),
    Cumulative = round(Cumulative * 100, 2)
  ) |>
  knitr::kable(
    col.names = c("PC", "Component", "Variance (%)", "Cumulative (%)")
  )
```

```{r}
#| label: variance-summary
#| echo: false
#| results: asis
v <- pca_fit$variance_df
cat(sprintf(
  "PC1 and PC2 explain **%.1f%%** and **%.1f%%** of variance (cumulative PC1–3: **%.1f%%**).",
  100 * v$Variance[1],
  100 * v$Variance[2],
  100 * v$Cumulative[min(3L, nrow(v))]
))
```

### Clusters vs study group

```{r}
#| label: cluster-table-k2
#| echo: false
knitr::kable(
  cluster_study_group_table(scores, "k2"),
  caption = "Participant counts: hierarchical clusters (k = 2) vs study group"
)
```

```{r}
#| label: cluster-table-k3
#| echo: false
knitr::kable(
  cluster_study_group_table(scores, "k3"),
  caption = "Participant counts: hierarchical clusters (k = 3) vs study group"
)
```

```{r}
#| label: cluster-interpretation
#| echo: false
#| results: asis
cat(
  "Clusters are **unsupervised** and need not align with clinical study group. ",
  "Compare tables and PCA colours to see whether diet–mycobiome structure tracks disease status or forms separate groupings.",
  sep = ""
)
```

## Figures

### Cumulative variance (scree)

```{r}
#| label: fig-variance
#| fig-cap: "Cumulative proportion of variance explained by principal components."
#| fig-width: 7
#| fig-height: 4.5
#| fig-align: center

plot_pca_variance(pca_fit$variance_df, max_pc = min(10L, n_participants))
```

### PCA scores (PC1 vs PC2)

```{r}
#| label: fig-pca-study-group
#| fig-cap: "PCA of combined nutrient and genus features. Points are participants; colour is study group."
#| fig-width: 8
#| fig-height: 5.5
#| fig-align: center

plot_pca_scores(
  scores,
  colour_col = "Study_group_new",
  variance_df = pca_fit$variance_df,
  title = NULL
)
```

### PCA with hierarchical clusters (k = 2)

```{r}
#| label: fig-pca-k2
#| fig-cap: "Same PCA space with point shape indicating Ward clusters (k = 2)."
#| fig-width: 8
#| fig-height: 5.5
#| fig-align: center

plot_pca_scores(
  scores,
  colour_col = "Study_group_new",
  shape_col = "k2",
  variance_df = pca_fit$variance_df,
  title = "PCA with hierarchical clusters (k = 2)"
)
```

### Top PC1 loadings

```{r}
#| label: fig-loadings
#| fig-cap: "Features with largest absolute loadings on PC1 (nutrients and genera)."
#| fig-width: 7
#| fig-height: 5
#| fig-align: center

plot_top_pca_loadings(pca_fit$loadings, pc = "PC1", n_top = 10L)
```

### Clustering dendrogram

```{r}
#| label: fig-dendrogram
#| fig-cap: "Ward hierarchical clustering dendrogram on scaled diet + genus features (participant IDs as leaves)."
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
#| warning: false

plot_cluster_dendrogram(cluster_fit)
```

```{r}
#| label: save-figures
#| include: false
dir.create(here("figures"), showWarnings = FALSE, recursive = TRUE)
ggsave(
  here("figures", "pca-diet-mycobiome-by-group.png"),
  plot_pca_scores(scores, "Study_group_new", variance_df = pca_fit$variance_df),
  width = 8,
  height = 5.5,
  dpi = 300
)
ggsave(
  here("figures", "pca-diet-mycobiome-variance.png"),
  plot_pca_variance(pca_fit$variance_df, max_pc = min(10L, n_participants)),
  width = 7,
  height = 4.5,
  dpi = 300
)
ggsave(
  here("figures", "pca-diet-mycobiome-loadings-pc1.png"),
  plot_top_pca_loadings(pca_fit$loadings, "PC1", 10L),
  width = 7,
  height = 5,
  dpi = 300
)
```

## Interpretation

PCA reduces the joint nutrient + genus feature space to orthogonal axes for visualization. If participants with similar diets and mycobiomes sit near each other, they may form visual groups; **formal cluster validity was not assessed** (silhouette, gap statistic, etc.) given small n.

Hierarchical clusters partition participants without using `Study_group_new`. Agreement between clusters and clinical group supports an integrated diet–mycobiome signal; disagreement suggests heterogeneity within disease labels or driven by diet/taxa not captured by group alone.

## Reproducibility

- **Helpers:** `stats/R/pca_clustering_helpers.R`, `stats/R/nutrient_association_helpers.R`, `stats/R/permanova_helpers.R`
- **Notebook:** `notebooks/pca_tsne.Rmd` (taxa-only PCA/t-SNE)
- **Render:** `quarto render stats` from repository root
