---
title: "PERMANOVA genus beta diversity by study group"
description: "Bray–Curtis dissimilarity on genus relative abundance; study-group effect tested with permutational MANOVA."
date: 2026-05-19
categories: [permanova, beta-diversity, genus, pre-specified]
author: "IBD Capstone team"
---

## Research question

Does **study group** explain differences in **genus-level** community composition?

**Null:** Multivariate centroids do not differ among study groups (PERMANOVA on Bray–Curtis distances).

## Data

- **Response:** Genus relative abundance (`g__*` columns after wide reshape).
- **Predictor:** `Study_group_new`.
- **Input:** `data/processed/genus.csv` (long format from `make save`; pivoted in this post).

## Methods

| Item | Choice |
|------|--------|
| Test | PERMANOVA (`vegan::adonis2`) |
| Distance | Bray–Curtis |
| Formula | `dist ~ Study_group_new` |
| Permutations | 999 |
| Visualization | PCoA (`cmdscale`, k = 2), 95% ellipses |

```{r}
#| label: setup
#| include: false
library(here)
library(knitr)
# Anchor here to stats/ so paths work both via `quarto render` and interactive IDE sessions.
here::i_am("posts/2026-05-19-permanova-genus-beta-diversity/index.qmd")
source(here::here("R", "permanova_helpers.R"))

level <- "genus"
cfg <- TAXA_LEVELS[[level]]
wide_df <- load_taxa_wide(level)
fit <- run_permanova_study_group(wide_df, cfg$prefix)
permanova_result <- fit$permanova
```

## Results

```{r}
#| label: permanova-table
#| echo: false
knitr::kable(
  permanova_to_df(permanova_result),
  digits = 4,
  col.names = c("Term", "Df", "Sum of squares", "R²", "F", "Pr(>F)")
)
```

```{r}
#| label: results-inline
#| echo: false
#| results: asis
cat(permanova_summary_text(permanova_result, cfg$label))
```

## Figure

```{r}
#| label: fig-pcoa
#| fig-cap: "PCoA of genus Bray–Curtis dissimilarity by study group."
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center

plot_pcoa_study_group(
  fit$dist,
  fit$metadata,
  permanova_result,
  cfg$label
)
```

```{r}
#| label: save-figure
#| include: false
dir.create(here("figures"), showWarnings = FALSE, recursive = TRUE)
ggsave(
  here("figures", "permanova-genus-pcoa.png"),
  width = 7,
  height = 5.5,
  dpi = 300
)
```

## Reproducibility

- **Helpers:** `stats/R/permanova_helpers.R`
- **Pipeline:** `make wrangle` → `make save`
