---
title: "Short title of the test"
description: "One sentence: what was tested and on which outcome."
date: 2026-05-19
categories: [exploratory]
author: "IBD Capstone Team"
freeze: true
---

## Research question

What hypothesis does this test address? State the null in plain language.

## Data

- **Outcome / response:**
- **Predictor(s):**
- **Unit of analysis:** (sample, participant, day, …)
- **N:** 
- **Input file(s):** `data/processed/...` or `data/intermediate/...`
- **Inclusion / exclusion:** (e.g. samples removed in `02_data_wrangling.R`)

## Methods

| Item | Choice |
|------|--------|
| Test | e.g. Kruskal–Wallis, PERMANOVA (`adonis2`) |
| R package | e.g. `vegan`, `rstatix` |
| Formula | `Value ~ Study_group_new` |
| Distance / link | (if applicable) Bray–Curtis |
| Permutations | (if applicable) 999 |
| Multiple testing | none / BH across metrics |
| Significance | α = 0.05 |

```{r}
#| label: setup
#| include: false
library(dplyr)
library(here)

# here::i_am() forces here to anchor to stats/ (where stats/.here lives) regardless of how
# the document is opened. Without it, interactive IDE sessions anchor to the .Rproj root
# Replace the path below with this post's directory name.
here::i_am("posts/<post-directory-name>/index.qmd")
# library(...)  # add packages used below
```

```{r}
#| label: run-test
# Load data and run the test. Keep parameters explicit for reproducibility.
# result <- ...
```

## Assumptions and diagnostics

- Independence:
- Exchangeability / rank-based:
- Other checks performed:

## Results

```{r}
#| label: results-table
#| echo: false
# knitr::kable(result, digits = 4)
```

**Summary in words:** (e.g. “Shannon diversity differed by study group, χ²(2) = …, p = …”.)

## Figure

```{r}
#| label: fig-main
#| fig-cap: "Caption describing what the plot shows."
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
# ggplot(...) + ...
```

Optional: save a static copy for slides or the manuscript.

```{r}
#| label: save-figure
#| eval: false
# ggsave(here("figures", "my-test.png"), width = 7, height = 5, dpi = 300)
```

## Interpretation

What can and cannot be concluded from this test?

## Reproducibility

- **Script / notebook:** `path/to/script.R` or `notebooks/....Rmd`
- **Rendered:** `quarto render stats` from repository root
- **Session:** see footer when `freeze: false` on first render
